$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-05 paper/paper.tex
Full pass

\todo{spellcheck}
\todo{make sure all citations are there}
\todo{rename gradual to something else...}

@@

Recent work has pushed the linear \picalc{} into the spotlight again \cite{KPT96}, thanks to an encoding of session types into linear types \cite{DardhaGS12,Dardha14,DardhaGS17}.
This encoding not only has theoretical benefits in terms of the expressivity of session types and the reusability of theoretical results from the linear \picalc{}, but it is useful at a practical level too.
The encoding has been used as a technique to implement session types in mainstream programming languages such as OCaml \cite{Padovani17} and Scala \cite{ScalasY16,ScalasDHY17}.
This allows us to use the linear \picalc{} as an underlying theoretical and practical framework on top of which session types can be built.

@@

\subsection{Contributions}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: \autoref{syntax} uses type level de Bruijn indices \cite{deBruijn1972} to introduce a syntax that is \emph{well-scoped by construction}: every free variable is accounted for in the type of the process that uses it.

@@

  \item \textbf{Leftover typing for the \picalc{} with linear, gradual and shared types}:
  \autoref{type-system} provides the type system for a resource-aware \picalc{}.
  \begin{itemize}
    \item \textbf{Algebras on multiplicities}: The user can provide \emph{resource-aware} algebras, which are then applied to the type system (\autoref{multiplicities}).
    Linear types, gradual types and shared types are all instances of the same resource-aware algebra, giving rise to three different type systems under the same framework.
    Any \emph{partial commutative monoid} that is \emph{functional}, \emph{cancellative} and has a minimal element is a valid such algebra.

@@

Data type definitions ($\type{\N}$) use double lines and index-free synonyms (\textsc{Nat}) as rule names (for ease of reference).
We otherwise use constructor names ($\constr{0}$ and $\suc$) to name our typing rules.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned by type definitions but omitted in constructors.

@@

A variable occurring under $n$ binders can refer to other $n$ possible variables.
References outside of that range are meaningless.
It is useful to rule out these ill-scoped nonsensical terms syntactically.
In \autoref{fig:syntax}, we do so by introducing the indexed family of types $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.
We index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
Every time we go under a binder, we increase the index of the continuation process, allowing the variable references within it to refer to one more thing.

@@

The first three rules ($\constr{comp}-$*) state associativity, symmetry, and $\PO$ as the neutral element of parallel composition, respectively.
The last three ($\constr{scope}-$*) state garbage collection, scope extrusion and commutativity of restrictions, respectively.

In $\constr{scope-ext}$ the type $\Unused_i \; Q$ is an inductive proof asserting that the variable index $i$ does not appear neither in the inputs nor in the outputs of $Q$.
The function $\func{lower}_i \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $i$.
In $\constr{scope-comm}$ the function $\func{swap}_i \; P$ traverses $P$ (of type $\Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all the above, $i$ is incremented every time we go under a binder.

@@

The operational semantics of the \picalc{} is defined by a reduction relation \textsc{Reduces} $\reduce{c}$ indexed by the channel $c$ on which communication occurs (\autoref{fig:reduction}).
We keep track of channel $c$ so we can state subject reduction (\autoref{subject-reduction}).
We distinguish between channels created inside a process ($\constr{internal}$) and channels created outside ($\constr{external}\ i$, where $i$ is the de Bruijn index of the variable).

In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel with index ${i}$.
As a result of that communication, the continuation of the input process $P$ has all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After substitution, $\subst{P}{\suc j}{\constr{0}}$ is \emph{lowered} --- all variable references are decreased by one (to do so, we apply \autoref{lm:subst-unused} in \autoref{app:lemmas} to obtain a proof $\Unused_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}})$).
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input nor output, as doing so would not preserve the sequencing of actions \cite{Sangio01}.

In $\constr{res}$, we use a $\func{dec}$ function to decrement the index of channel $c$ as we wrap processes $P$ and $Q$ inside a binder.
This $\func{dec}$ function saturates at $\constr{internal}$: wrapping an internally reducing process with a binder will result in an internally reducing process.

@@

Linear, gradual and shared types are all defined as an instance of our usage algebra.
Their use in typing derivations is illustrated in \autoref{example-derivations}.
\begin{description}
  \item [Linear]
    The carrier is a type with two trivial elements $\constr{zero}$ and $\constr{one}$.
    The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{one}$), and is uninhabited otherwise.
    All the properties of the algebra follow trivially.
    As an example, the multiplicity pair $\ell_i$ is represented as $(\constr{one} \comma \constr{zero})$.

@@

We keep typing contexts ($\PreCtx$, in \autoref{fig:prectx}) and usage contexts ($\Ctx$ in \autoref{fig:ctx}) separate.
The former are preserved throughout a typing derivation; the latter are transformed as a result of input, output, and context splits.

\begin{nidefinition}[Types and length-indexed typing contexts]
  A \emph{type} is either a unit type ($\unit$), a base type ($\base{n}$) or a channel type ($\channel{t}{x}$) defined in \autoref{fig:prectx}.

  A \emph{length-indexed typing context} $\PreCtx_n$, is defined by the last row of rules in \autoref{fig:prectx}.
  A typing context is either empty $[]$ or the result of appending a type $t$ to an existing context$\gamma$.
  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\Type : \Set}
    \; \textsc{Type}

@@

The unit type $\unit$ serves as a proof of inhabitance for types.
The base type $\base{n}$ uses natural numbers as placeholders for types (the host language can then interpret the former into the latter --- e.g. $0$ as booleans, $1$ as natural numbers) so that we avoid having to deal with universe polymorphism.
The type $\channel{t}{x}$ of a channel determines what type of data ${t}$ and what usage annotations ${x}$ are sent over that channel.

Notice that our notation of channel types is similar to Kobayashi's $[t]\mathbf{chan}_{(i^{y},o^{z})}$, where $y,z$ are the multiplicities of input and output, respectively \cite{K07}.

@@

    \inferrule
        {\Gamma : \Ctx_{idxs} \\ x : \Usage_{idx} ^{\func{2}}}
        {\Gamma \comma x : \Ctx_{idxs \comma idx}}
  \end{mathpar}
  \caption{Length-indexed context of carrier indices with a context of multiplicities on top.}
  \label{fig:ctx}
  \end{figure}
\end{nidefinition}

\begin{note}
  We lift the monoidal operation on usage annotations $\opsquared{x}{y}{z}$ to a monoidal operation $\opctx{\Gamma}{\Delta}{\Xi}$ on usage contexts that have a common underlying context of indices.
  All the properties of the partial monoid are trivially lifted.
\end{note}

@@

We use \emph{leftover typing} \cite{Allais2018a} for our type system, an approach that, in addition to the usual typing context $\PreCtx_n$ and (input) usage context $\Ctx_{idxs}$, adds an extra \emph{output usage context} to the typing rules.

This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
Leftover typing presents three main advantages:
\begin{itemize}
  \item \textbf{Encapsulation}:
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgement with both an input and an output usage context.
  \item \textbf{No context split}:
    No need for an extensional context splitting proof for parallel composition.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item \textbf{New theorems}:
    For the first time, theorems like \textbf{framing} (\autoref{thm:framing}), \textbf{weakening} (\autoref{thm:weakening}) and \textbf{strengthening} (\autoref{thm:strengthening}) can be stated within the linear \picalc{} thanks to the presence of a leftover context.
\end{itemize}

@@

The typing judgement for variables $\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta}$ asserts that ``index $i$ is of type $t$ in typing context $\gamma$, and subtracting $y$ at position $i$ from input usage context $\Gamma$ results in leftover $\Delta$''.
The typing judgement for processes $\types{\gamma}{\Gamma}{P}{\Delta}$ asserts that ``process $P$ is well-typed under typing context $\gamma$, usage input context $\Gamma$ and leftovers $\Delta$''.

@@

\begin{example}[Channels]
  Let us illustrate the use of channels with an example.
  \begin{lstlisting}[label=lst:channels,mathescape,numbers=left,caption=The underscore \_ introduces an anonymous declaration immediately followed by its definition.]
  _ : $\contains{\constr{[]} \comma \channel{\unit}{\li} \comma \unit} {\constr{[]} \comma \lio \comma \lio} {\suc \; \constr{0}} {\channel{\unit}{\li}} {\li} {\constr{[]} \comma \lo \comma \lio}$
  _ = $\suc \; \constr{0}$
  \end{lstlisting}
  In \autoref{lst:channels}, the constructor $\suc$ steps under the outermost variable in the context, preserving its usage annotation $\lio$ from input to output.
  The constructor $\constr{0}$ asserts that the next variable is of type $\channel{\unit}{\li}$, and that we can split its input annotation $\lio$, taking away $\li$ and leaving leftovers $\lo$.
\end{example}

@@

  The inaction process in rule $\constr{end}$ does not change usage annotations.
  The scope restriction in rule $\constr{chan}$ expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ that is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that these are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $(y \comma y)$, which it must completely exhaust.

@@

  Similarly to input, the output process in rule $\constr{send}$ requires a channel $chan_i$ at index $i$ with usage $\lo$ available, such that data with type $t$ and usage $x$ can be sent over it.
  We use the leftover context $\Delta$ to type the continuation process, which needs an element $loc_j$ at index $j$ with type $t$ and usage $x$, as per the type of channel $chan_i$.
  The leftovers $\Xi$ are used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.

  Parallel composition in rule $\constr{comp}$ uses the leftovers of the left-hand process to type the right-hand process.
  By keeping track in the typing derivation of $P$ the resources it uses, we save the user from having to provide an extrinsic proof of context split.
\end{nidefinition}

\begin{example}[Typing derivation]
  \label{example-derivations}
  We will now illustrate a typing derivation of the process $\func{p}$ using linear and shared types (Listing \autoref{lst:derivation}).

  Let $\func{\omega}$ be a shortcut for $(\one \comma \one)$ in an algebra of shared types.
  Let $\epsilon$ be the empty usage context --- where all usage annotations are $(\zero \comma \zero)$.
  To type $\func{p}$, we need a channel over which we can transmit another channel.
  We therefore create a new channel with usage annotations $(\one \comma \one)$ capable of transmitting data of type $\channel{\unit}{\func{\omega}}$, and declare (with $\li$) that we will only transmit its input capability (line 7).
  The left-hand process is typed following its syntax (line 8).
  On the right-hand we need the channel that is going to be sent (line 9).
  The type $\unit$ of this channel is entirely determined by the type of the channel over which it is going to be transmitted.
  Its usage annotation is $(\one \comma \one)$: we will transmit $\li$ and save $\lo$ for the continuation process.

@@

Removing an unused variable preserves the well-typedness of a process.
Let $\func{del}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
\begin{nitheorem}[Strengthening]
  \label{thm:strengthening} 
  Let $P$ be a well-typed process in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable index not in $P$, such that $\Unused_i \; P$.
  Then lowering every variable greater than or equal to $i$ in $P$ is well-typed in $\types{\func{del}_i \; \gamma}{\func{del}_i \; \Gamma}{\func{lower}_i \; P}{\func{del}_i \; \Xi}$.
\end{nitheorem}

\begin{remark}
  It is possible for the usage annotations of a variable $i$ within a process $P$ to be preserved as leftovers, yet for the variable $i$ to be mentioned within $P$ --- $P$ can choose to send $\lz$ usage annotations of $i$ over a channel.
\end{remark}

Swapping two variables preserves the well-typedness of a process.
Let $\func{sw}_i$ be a function that swaps variables $i$ and $\suc i$.
For simplicity, we overload $\func{sw}$ and use it to swap types in a typing context --- we swap variable references in a process with $swap_i \; P$, as first used in structural congruence (\autoref{structural-congruence}).
\begin{nitheorem}[Swapping]
  \label{thm:swapping}
  Let $P$ be a well-typed process in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{sw}_i \; \gamma}{\func{sw}_i \; \Gamma}{\func{swap}_i \; P}{\func{sw}_i \; \Xi}$.
\end{nitheorem}

\begin{remark}
  We limit ourselves to the swapping of adjacent variables $i$ and $\suc i$.
  Doing so is enough to prove subject reduction.
  More generally, any two variables can be swapped by performing a series of adjacent variable swappings.
\end{remark}

\begin{proof}
  All the above theorems are proved by induction on \textsc{Process} and \textsc{Var}.
  For details, refer to the mechanisation in Agda \cite{Zalakain2020Agda}.
\end{proof}

@@

Subject congruence theorem states that applying structural congruence rules to a well-typed process preserves its well-typedness.
\begin{nitheorem}[Subject congruence]
  \label{thm:subject-congruence}
  If $P \eq{} Q$ and $\types{\gamma}{\Gamma}{P}{\Xi}$, then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{nitheorem}

\begin{proof}[Proof (Sketch)]
  The proof is induction on \textsc{Equals} $\eq{}$.
  Here we will only consider the base cases for $\constr{struct}$ (and their symmetric variants under $\constr{sym}$), which are nontrivial. Full proof in \cite{Zalakain2020Agda}.
  We proceed by induction on \textsc{StructCong} $\equiv$:
  \begin{itemize}
    \item
      Case $\constr{comp-assoc}$: trivial, as leftover typing is naturally associative.
    \item
      Case $\constr{comp-sym}$ for $\comp{P}{Q}$: we use framing (\autoref{thm:framing}) to shift the output context of $P$ to the one of $Q$; and the input context of $Q$ to the one of $P$.
      The point at which $Q$ and $P$ join in $\comp{Q}{P}$ will no longer be the same.
    \item
      Case $\constr{comp-end}$: trivial, as the typing rule for $\PO$ has the same input and output contexts.
    \item
      Case $\constr{scope-end}$: we show that the usage annotation of the newly created channel must be $\lz$, making the proof trivial.
      In the opposite direction, we instantiate the newly created channel to a type $\unit$ and a usage annotation $\lz$..
    \item
      Case $\constr{scope-ext}$ for $\new \comp{P}{Q}$: we use \autoref{lm:types-unused} to show that $P$ preserves the usage annotations of the unused variable.
      We then use strengthening (\autoref{thm:strengthening}) to delete the unused variable from $P$, and compose the result with typing derivation on $Q$ wrapped in $\constr{chan}$.
      In the reverse direction, we weaken (\autoref{thm:weakening}) $P$ so that it accepts the additional variable.
      We then use \autoref{lm:lower-lift} to show that lowering and then lifting $P$ results in $P$.
    \item
      Case $\constr{scope-comm}$: in one direction we use swapping (\autoref{thm:swapping}).
      In the reverse direction, we use swapping and then \autoref{lm:swap-swap} to show that swapping two elements in $P$ twice leaves $P$ unchanged.
  \end{itemize}
\end{proof}

@@

\begin{nitheorem}[Substitution]
  \label{thm:substitution}
  Let $P$ be a process typed under $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \lz}$.
  Let $\ni_j m$ be a \textsc{VarRef} $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute the variable references to $\constr{0}$ in $P$ with $\suc j$ so that the result is well-typed under $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{0}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}
  Use framing to derive $\contains{\gamma}{\Gamma}{j}{m}{\Theta}$ and $\types{\gamma \comma t}{\Theta \comma m}{P}{\Xi \comma \lz}$ for some $\Theta$.
  Supply these to the generalised \autoref{thm:subst-generalization}.
\end{proof}

>>>>

\todo{spellcheck}
\todo{make sure all citations are there}
\todo{rename gradual to something else...}
\todo{lemmas in appendix are referred to from here}

@@

Recent work has pushed the linear \picalc{} into the spotlight again \cite{KPT96} thanks to an encoding of session types into linear types \cite{DardhaGS12,Dardha14,DardhaGS17}.
This encoding not only has theoretical benefits in terms of the expressivity of session types and the reusability of theoretical results from the linear \picalc{}, but is useful at a practical level too.
The encoding has been used as a technique to implement session types in mainstream programming languages such as OCaml \cite{Padovani17} and Scala \cite{ScalasY16,ScalasDHY17}.
This allows us to use the linear \picalc{} as an underlying theoretical and practical framework on top of which session types can be built.

@@

\subsection{Contributions}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: \autoref{syntax} uses type level de Bruijn indices \cite{deBruijn1972, Dybjer1994} to introduce a syntax that is \emph{well-scoped by construction}: every free variable is accounted for in the type of the process that uses it.

@@

  \item \textbf{Leftover typing for the \picalc{} with linear, gradual and shared types}:
  \autoref{type-system} provides the type system for a resource-aware \picalc{}.
  \begin{itemize}
    \item \textbf{Algebras on multiplicities}: The user can provide \emph{resource-aware} algebras, which are then applied to the type system (\autoref{multiplicities}).
    Linear types, gradual types and shared types are all instances of the same resource-aware algebra, giving rise to three different type systems under the same framework.
    Any \emph{partial commutative monoid} that is \emph{functional}, \emph{cancellative} and has a \emph{minimal element} is a valid such algebra.

@@

Data type definitions ($\type{\N}$) use double lines and index-free synonyms (\textsc{Nat}) as rule names (for ease of reference).
We otherwise use constructor names ($\constr{0}$ and $\suc$) to name our typing rules.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned by type definitions but omitted by constructors.

@@

A variable reference occurring under $n$ binders can refer to $n$ distinct variables.
References outside of that range are meaningless.
It is useful to rule out these ill-scoped nonsensical terms syntactically.
In \autoref{fig:syntax}, we do so by introducing the indexed family of types \cite{Dybjer1994} $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.
We index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
Every time we go under a binder, we increase the index of the continuation process, allowing the variable references within to refer to one more thing.

@@

The first three rules ($\constr{comp}-$*) state associativity, symmetry, and $\PO$ as being the neutral element of parallel composition, respectively.
The last three ($\constr{scope}-$*) state garbage collection, scope extrusion and commutativity of restrictions, respectively.

In $\constr{scope-ext}$ the type $\Unused_i \; Q$ is an inductive proof asserting that the variable index $i$ does not appear neither in the inputs nor in the outputs of $Q$.
The function $\func{lower}_i \; Q \; uQ$ traverses $Q$ decrementing every index greater than $i$.
In $\constr{scope-comm}$ the function $\func{swap}_i \; P$ traverses $P$ (of type $\Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all the above, $i$ is incremented every time we go under a binder.

@@

The operational semantics of the \picalc{} is defined as a reduction relation \textsc{Reduces} $\reduce{c}$ indexed by the channel $c$ on which communication occurs (\autoref{fig:reduction}).
We keep track of channel $c$ so we can state subject reduction (\autoref{subject-reduction}).
We distinguish between channels that are created inside of the process ($\constr{internal}$) and channels that are created outside ($\constr{external}\ i$, where $i$ is the index of the channel variable).

In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel with index ${i}$.
As a result of that communication, the continuation of the input process $P$ has all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After substitution, $\subst{P}{\suc j}{\constr{0}}$ is \emph{lowered} --- all variable references are decreased by one (we apply \autoref{lm:subst-unused} in \autoref{app:lemmas} to obtain a proof $\Unused_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}})$).
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input nor output, as doing so would not preserve the sequencing of actions \cite{Sangio01}.

In $\constr{res}$, we use $\func{dec}$ to decrement the index of channel $c$ as we wrap processes $P$ and $Q$ inside a binder.
This $\func{dec}$ function saturates at $\constr{internal}$: wrapping an internally reducing process with a binder will result in an internally reducing process.

@@

Linear, gradual and shared types are all defined as an instance of our usage algebra.
Their use in typing derivations is illustrated in \autoref{example-derivations}.
\begin{description}
  \item [Linear]
    The carrier is a type with two trivial constructors $\constr{zero}$ and $\constr{one}$.
    The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{one}$), and is uninhabited otherwise.
    All the properties of the algebra follow trivially.
    As an example, the multiplicity pair $\ell_i$ is represented as $(\constr{one} \comma \constr{zero})$.

@@

We keep typing contexts ($\PreCtx$, in \autoref{fig:prectx}) and usage contexts ($\Ctx$ in \autoref{fig:ctx}) separate.
The former are preserved throughout typing derivations; the latter are transformed as a result of input, output, and context splits.

\begin{nidefinition}[Types and typing contexts]
  A \emph{type} is either a unit type ($\unit$), a base type ($\base{n}$) or a channel type ($\channel{t}{x}$) (\autoref{fig:prectx}).

  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\Type : \Set}
    \; \textsc{Type}

@@

The unit type $\unit$ serves as a proof of inhabitance for types.
The base type $\base{n}$ uses natural numbers as placeholders for types (the host language can then interpret the former into the latter --- e.g. $0$ as booleans, $1$ as natural numbers) so that we avoid having to deal with universe polymorphism.
The type $\channel{t}{x}$ of a channel determines what type of data ${t}$ and what usage annotations ${x}$ are sent over that channel.
Notice that this notation is similar to Kobayashi's $[t]\mathbf{chan}_{(i^{y},o^{z})}$, where $y,z$ are the multiplicities of input and output, respectively \cite{K07}.

A \emph{typing context} (last row in \autoref{fig:prectx}) is a length-indexed list of types, and is either empty $[]$ or the result of appending a type $t$ to an existing context $\gamma$.

@@

    \inferrule
        {\Gamma : \Ctx_{idxs} \\ x : \Usage_{idx} ^{\func{2}}}
        {\Gamma \comma x : \Ctx_{idxs \comma idx}}
  \end{mathpar}
  \caption{Length-indexed context of carrier indices with a context of usage annotations on top.}
  \label{fig:ctx}
  \end{figure}
\end{nidefinition}

\begin{note}
  We lift the monoidal operation on usage annotations $\opsquared{x}{y}{z}$ to a monoidal operation $\opctx{\Gamma}{\Delta}{\Xi}$ on usage contexts that have a common underlying context of indices.
  All the properties of the partial monoid are lifted element-wise.
\end{note}

@@

We use \emph{leftover typing} \cite{Allais2018a} for our type system, an approach that, in addition to the usual typing context $\PreCtx_n$ and (input) usage context $\Ctx_{idxs}$, adds an extra \emph{output usage context} $\Ctx_{idxs}$ to the typing rules.

This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
Leftover typing presents three main advantages:
\begin{itemize}
  \item \textbf{Encapsulation}:
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgement with both an input and an output usage context.
  \item \textbf{No extensional context split}:
    No need for an extensional context splitting proof for parallel composition.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item \textbf{New theorems}:
    For the first time, theorems like \textbf{framing} (\autoref{thm:framing}), \textbf{weakening} (\autoref{thm:weakening}) and \textbf{strengthening} (\autoref{thm:strengthening}) can be stated within the linear \picalc{} thanks to the presence of a leftover context.
\end{itemize}

@@

The typing judgement for variables $\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta}$ asserts that ``index $i$ is of type $t$ in typing context $\gamma$, and subtracting $y$ at position $i$ from input usage context $\Gamma$ results in leftovers $\Delta$''.
The typing judgement for processes $\types{\gamma}{\Gamma}{P}{\Delta}$ asserts that ``process $P$ is well-typed under typing context $\gamma$, usage input context $\Gamma$ and leftovers $\Delta$''.

@@

\begin{example}[Variable reference]
  Let us illustrate the use of variable reference relations with an example.
  \begin{lstlisting}[label=lst:channels,mathescape,numbers=left,caption=The underscore \_ introduces an anonymous declaration immediately followed by its definition.]
  _ : $\contains{\constr{[]} \comma \channel{\unit}{\li} \comma \unit} {\constr{[]} \comma \lio \comma \lio} {\suc \; \constr{0}} {\channel{\unit}{\li}} {\li} {\constr{[]} \comma \lo \comma \lio}$
  _ = $\suc \; \constr{0}$
  \end{lstlisting}
  In \autoref{lst:channels}, the constructor $\suc$ steps under the outermost variable in the context, preserving its usage annotation $\lio$ from input to output.
  The constructor $\constr{0}$ asserts that the next variable is of type $\channel{\unit}{\li}$, and that we can split its input annotation $\lio$, taking away $\li$ and leaving $\lo$ as the leftover.
\end{example}

@@

  The inaction process in rule $\constr{end}$ does not change usage annotations.
  The scope restriction in rule $\constr{chan}$ expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ of what is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that they are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $(y \comma y)$, which it must completely exhaust.

@@

  Similarly to input, the output process in rule $\constr{send}$ requires a channel $chan_i$ at index $i$ with usage $\lo$ available, such that data with type $t$ and usage $x$ can be sent over it.
  We use the leftover context $\Delta$ to type the continuation process, which needs an element $loc_j$ at index $j$ with type $t$ and usage $x$, as per the type of the channel $chan_i$.
  The leftovers $\Xi$ are used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.

  Parallel composition in rule $\constr{comp}$ uses the leftovers of the left-hand process to type the right-hand process.
  By keeping track in the typing derivation of $P$ of the resources $P$ uses, we can use them to type $Q$ and save the user from having to provide an extrinsic proof of context split.
\end{nidefinition}

\begin{example}[Typing derivation]
  \label{example-derivations}
  We will now illustrate a typing derivation of the process $\func{p}$ using linear and shared types (\autoref{lst:derivation}).

  Let $\func{\omega}$ be a shortcut for $(\one \comma \one)$ in an algebra of shared types.
  Let $\epsilon$ be the empty usage context --- where all usage annotations are $(\zero \comma \zero)$.
  To type $\func{p}$, we need a channel over which we can transmit another channel.
  We therefore create a new channel with usage annotations $(\one \comma \one)$ capable of transmitting data of type $\channel{\unit}{\func{\omega}}$, and declare (with $\li$) that we only transmit input capabilities over it (line 7).
  The left-hand process is typed following its syntax (line 8).
  On the right-hand we need the channel that is going to be sent (line 9).
  The type $\unit$ of this channel is entirely determined by the type $\channel{\unit}{\func{\omega}}$ of the channel over which it is going to be transmitted.
  Its usage annotation is $(\one \comma \one)$: we will transmit $(\one \comma \zero)$ and save $(\zero \comma \one)$ for the continuation process.

@@

Removing an unused variable preserves the well-typedness of a process.
Let $\func{del}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
\begin{nitheorem}[Strengthening]
  \label{thm:strengthening} 
  Let $P$ be a well-typed process in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable index not in $P$, such that $\Unused_i \; P$.
  Then lowering every variable greater than $i$ in $P$ is well-typed in $\types{\func{del}_i \; \gamma}{\func{del}_i \; \Gamma}{\func{lower}_i \; P}{\func{del}_i \; \Xi}$.
\end{nitheorem}

\begin{remark}
  It is possible for the usage annotations of a variable $i$ within a process $P$ to be preserved as leftovers, yet for the variable $i$ to be mentioned by $P$: $P$ can choose to send $\lz$ usage annotations of $i$ over a channel.
\end{remark}

Swapping two variables preserves the well-typedness of a process.
Let $\func{sw}_i$ be a function that swaps variables $i$ and $\suc i$.
For simplicity, we overload $\func{sw}$ and use it to swap types in typing contexts and usage annotations in usage contexts. We use $\func{swap}_i \; P$ to swap variables $i$ and $\suc i$ within a process $P$, as introduced in structural congruence (\autoref{structural-congruence}).
\begin{nitheorem}[Swapping]
  \label{thm:swapping}
  Let $P$ be a well-typed process in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{sw}_i \; \gamma}{\func{sw}_i \; \Gamma}{\func{swap}_i \; P}{\func{sw}_i \; \Xi}$.
\end{nitheorem}

\begin{remark}
  We limit ourselves to the swapping of adjacent variables $i$ and $\suc i$.
  Doing so is enough to prove subject reduction.
  More generally, any two variables can be swapped by a series of adjacent variable swappings.
\end{remark}

\begin{proof}[Proof (Sketch)]
  All the above theorems are proved by induction on \textsc{Process} and \textsc{Var}.
  For details, refer to the mechanisation in Agda \cite{Zalakain2020Agda}.
\end{proof}

@@

Subject congruence states that applying structural congruence rules to a well-typed process preserves its well-typedness.
\begin{nitheorem}[Subject congruence]
  \label{thm:subject-congruence}
  If $P \eq{} Q$ and $\types{\gamma}{\Gamma}{P}{\Xi}$, then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{nitheorem}

\begin{proof}[Proof (Sketch)]
  The proof is by induction on \textsc{Equals} $\eq{}$.
  Here we will only consider the base cases for $\constr{struct}$ (and their symmetric variants under $\constr{sym}$), which are nontrivial. Full proof in \cite{Zalakain2020Agda}.
  We proceed by induction on \textsc{StructCong} $\eqeq$:
  \begin{itemize}
    \item
      Case $\constr{comp-assoc}$: trivial, as leftover typing is naturally associative.
    \item
      Case $\constr{comp-sym}$ for $\comp{P}{Q}$: we use framing (\autoref{thm:framing}) to shift the output context of $P$ to the one of $Q$; and the input context of $Q$ to the one of $P$.
      The point at which $Q$ and $P$ join in $\comp{Q}{P}$ will no longer be the same.
    \item
      Case $\constr{comp-end}$: trivial, as the typing rule for $\PO$ has the same input and output contexts.
    \item
      Case $\constr{scope-end}$: we show that the usage annotation of the newly created channel must be $\lz$, making the proof trivial.
      In the opposite direction, we instantiate the newly created channel to a type $\unit$ and a usage annotation $\lz$..
    \item
      Case $\constr{scope-ext}$ for $\new \comp{P}{Q}$: we use \autoref{lm:types-unused} to show that $P$ preserves the usage annotations of the unused variable.
      We then use strengthening (\autoref{thm:strengthening}) to delete the unused variable from $P$, and compose the result with the typing derivation on $Q$ wrapped in $\constr{chan}$.
      In the reverse direction, we weaken (\autoref{thm:weakening}) $P$ so that it accepts the additional variable.
      We then use \autoref{lm:lower-lift} to show that lowering and then lifting $P$ results in $P$.
    \item
      Case $\constr{scope-comm}$: in one direction we use swapping (\autoref{thm:swapping}).
      In the reverse direction, we use swapping and then \autoref{lm:swap-swap} to show that swapping two elements in $P$ twice leaves $P$ unchanged.
  \end{itemize}
\end{proof}

@@

\begin{nitheorem}[Substitution]
  \label{thm:substitution}
  Let $P$ be a well-typed in $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \lz}$.
  Let $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute the variable references to $\constr{0}$ in $P$ with $\suc j$ so that the result is well-typed in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{0}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}[Proof (Sketch)]
  Use framing to derive $\contains{\gamma}{\Gamma}{j}{t}{m}{\Theta}$ and $\types{\gamma \comma t}{\Theta \comma m}{P}{\Xi \comma \lz}$ for some $\Theta$.
  Use these to apply \autoref{thm:subst-generalization}.
\end{proof}
